The Pair Distribution Function of the Inhomogeneous Hard-Sphere Fluid

### 2014-03-10 papers/pair-correlation/paper.tex
modified text in paper.  Passing hot potato back Tested on crystal. Took 1:11.02.

%\newcommand{\derivation}[1]{#1} % Use this to show all derivations in detail
\newcommand{\derivation}[1]{} % Use this for nice pegagogical paper...
\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\jeffsays}[1]{{\color{red} [\green{Jeff:} \emph{#1}]}}
\newcommand{\pahosays}[1]{{\color{red} [\green{Paho:} \emph{#1}]}}

@@

\section{Our approach}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) +
    g(|r_{12}|, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
This approximation is constructed to reproduce the exact value for the
integral$ \int n(\rr_1)n(\rr_2)
g^{(2)}(\rr_1,\rr_2)\delta(|\rr_1-\rr_2|-\sigma)d\rr_1d\rr_2$.  This
approximation (described in Eq.~\ref{eq:g2-our-mean}) requires the
radial distribution function as a function of $r$ and $g_\sigma$.  One
approach is to compute $\eta$ from $g_\sigma$ and then find
$g(r,\eta)$ either using some approximate functional form
(e.g. Percus-Yevick) or using Monte Carlo simulation results.  In this
paper we use the latter (also for evaluating the approximations of
earlier works), and refer to this approximation (with an exact
$g(r,g_\sigma)$) as ``this work MC.''  \fixme{We should develop
  separate acroyms for each method.}

@@

%% The finalized form of our pair correlation function is:
%% \begin{align}
%%   g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
%% \end{align}
\pahosays{Was plotting the radial distribution function meant to be commented?}
%% \begin{figure}
%%   \centering
%%   \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
%%   %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
%%   \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
%% \end{figure}

In contrast to the previous works, which are based on the homogeneous
radial distribution function $g(r,\eta)$ expressed as a function of
distance and packing fraction, we have constructed our separable
approximation for the radial distribution function as $g(r,
g_\sigma)$, which is a function of the radius $r$ and its value at
contact $g_\sigma$.  One can convert between these two representation
using the Carnahan-Starling formula for $g_\sigma(\eta)$
\begin{equation}
  g_\sigma = \frac{1-\tfrac{\eta}{2}}{(1-\eta)^3}.
\end{equation}
The form $g(r, g_\sigma)$ is more convenient for treating the
inhomogeneous case, because we can directly compute the averaged value
of $g_\sigma$ for an inhomogeneous system~\cite{schulte2012using}.

We construct $g(r, g_\sigma)$ to satisfy several constraints.  Naturally,
\begin{equation}
  g(\sigma, g_\sigma) = g_\sigma.
\end{equation}
A second constraint applies to the integral of the total correlation
function
\begin{align}
  1 + n\int h(r)d\rr &= nkT\chi_T \label{eq:total-constraint}
\end{align}
where $h(r) = g(r) - 1$ is the total correlation function.
A final constraint is that the slope of the radial distribution
function at contact is given to a very good approximation by
\begin{align}
  g'(\sigma) \approx - h_\sigma g_\sigma
\end{align}
where $h_\sigma = g_\sigma - 1$ is the total correlation function.
Under these three constraints, we applied a least-squares fit to
simulation data for the radial distribution function at packing
fractions from 0.05 to 0.4 by steps of 0.05.  Our fitted radial
distribution function is given by
\begin{multline}
  g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
\begin{equation}
  I(g_\sigma) = \kappaZ^5 \frac{
    \frac{\chi-1}{24\eta} - h_\sigma \frac{(\kappaO-2
      g_\sigma)(\kappaI^2 + 4 \kappaI + 6)}{\kappaI^4}
    - h_\sigma\frac{\kappaO^2 + 2\kappaO + 2}{\kappaO}
  }{2 \kappaZ^2 + 12 \kappaZ + 24}
\end{equation}
\begin{equation}
  \chi = \frac{(1-\eta)^4}{\eta^4 - 4\eta^3 + 4\eta^2 + 4\eta + 1}
\end{equation}
\newcommand\nastyetacuberoot{\sqrt[3]{54 g_\sigma^2 -
    6\sqrt{81g_\sigma^4 - 6g_\sigma^3}}}
\begin{equation}
  \eta = 1 - \frac{1}{\nastyetacuberoot} - \frac{\nastyetacuberoot}{6g_\sigma}
\end{equation}
This function includes three fitted parameters, $\kappaO =
\kappazero$, $\kappaI = \kappaone$,
and $\kappaZ = \kappatwo$.
The fit is displayed in Fig.~\ref{fig:radial-distribution}.  In that
figure, it is apparent that the integral f $h(r)$ from
Eq.~\ref{eq:total-constraint} results in cancelling errors at large
distances.  The maximum error in $g(r)$ at packing fractions up to
0.4 is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr$.

\newcommand\paircaption{The pair distribution function near a hard wall, with
  packing fractions of 0.1 and 0.3 and $\rr_1$ in contact with the hard wall.
  On the left is a 2D plot of $g^{(2)}(\rr_1,\rr_2)$ as $\rr_2$ varies.  The top
  half of this figure shows the result of the Monte Carlo simulation, while the
  bottom half shows the result of the appproximation developed in this work.  On
  the right is a plot of $g^{(2)}(\rr_1,\rr_2)$ on the path illustrated in the
  figure to the left.  This plot compares the Monte Carlo results (black) with
  those of this work (blue), Sokolowski and Fischer
  (red)~\cite{sokolowski1992role}, and Fischer and Methfessel
  (green)~\cite{fischer1980born}.  The latter is only plotted at contact, where
  it is defined.  }

@@

We begin by examining the pair distribution function near a hard wall,
with a focus on the case where one of the two spheres is in contact
with the hard wall.  Figure~\ref{fig:pair-distribution}a
and~\ref{fig:pair-distribution}a compare the results of our
approximate functional with Monte Carlo simulation at packing
fractions of 0.1 and 0.3 respectively.  We see reasonable agreement at
the lower density, with some reduced oscillations at larger distances,
and a flatter angular dependence when the two spheres are in contact.
At the higher density, we see significant structure developing in the
simulation results that is not reflected in our approximation.

Figure~\ref{fig:pair-distribution}b
and~\ref{fig:pair-distribution}b show the pair distribution function
as plotted along a path illustrated in
Fig.~\ref{fig:pair-distribution}a.  These plots compare this work
with Monte Carlo results, as well as the approximations of Sokolowski
and Fischer~\cite{sokolowski1992role} and that of Fischer and
Methfessel~\cite{fischer1980born} at packing fractions of 0.1 and 0.3.
The approach of Fischer and Methfessel is only defined when the two
spheres are in contact, and is therefore only plotted on that segment
of the path.  As an input to both previous approximations we use the
hard sphere radial distribution function found with Monte Carlo
simulation, interpolated as necessary.  We find that both previous
approximations to the pair distribution function predict the
qualtitative angular dependence of the pair distribution function at
contact better than this work.  However, in each case the pair
distribution function has a systematic error at contact---either too
high or too low.  Thus while our approximation is smoother than either
of the existing approaches, its errors will have a tendancy to cancel
when used in a perturbation expansion.  At higher densities, the
approximation of Fischer and Methfessel requires evaluating the radial
distribution function at densities significantly higher than the
freezing density, which poses numerical difficulties when using the
radial distribution function from simulation.  When the two points are
more than a radius away from contact, we find that any of these
approaches gives a reasonable prediction, with our approach
underestimating the oscillations in $g^{(2)}$, as expected based on the
fit in Fig.~\ref{fig:radial-distribution}.

@@

\begin{figure*}
  \includegraphics[width=\textwidth]{figs/triplet-correlation-pretty-contact-3.pdf}\\
  \includegraphics[width=\textwidth]{figs/triplet-correlation-pretty-inbetween-3.pdf}
  \caption{\fixme{Make note of the double curves}\fixme{Add backward curve for
      this work mc?}\pahosays{would annotations be clearer if they were on the
      x/z ticks instead of arrows?}The triplet distribution function
    $g^{(3)}(\rr_1,\rr_2,\rr_3)$ at packing fraction 0.3, plotted when $\rr_1$
    and $\rr_2$ are in contact.  The color plot on the left shows the triplet
    distribution function, with white corresponding to the contact value of the
    \emph{pair} distribution function.  The plot on the right shows the value of
    $g^{(3)}$ along the path shown as a dotted line on the plot to the
    left.}\label{fig:triplet-contact-distribution}
\end{figure*}

@@

A particularly relevant quantitative test of a pair distribution
function is how well it predicts the interaction energy due to a pair
potential.  To this end, we have computed the error in the first term
in a high-temperature perturbation expansion $a_1$ for several pair
potentials.  In order to focus on effects at the interface, we have
defined a position-dependent pair interaction energy as
\begin{align}
  \frac{dF_1}{dz} &=
  \tfrac12 \int g^{(2)}_{HS}(\rr,\rr')n(\rr)n(\rr')\Phi(|\rr-\rr'|)
  d\rr'\, dxdy\label{eq:da1}
\end{align}
which gives the contribution to the mean-field free energy due to
molecules located at position $z$.

>>>>

%\newcommand{\derivation}[1]{#1} % Use this to show all derivations in detail
\newcommand{\derivation}[1]{} % Use this for nice pegagogical paper...
\newcommand{\davidsays}[1]{{\color{red} [\green{David:} \emph{#1}]}}
\newcommand{\jeffsays}[1]{{\color{red} [\green{Jeff:} \emph{#1}]}}

@@

\section{Our approach}
%% We approximate the pair correlation with the average of two radial
%% distribution functions, evaluated at the distance between the two
%% points, that are themselves functions of the averaged pair correlation
%% at contact $g_{\sigma}(\rr)$ discussed above~\cite{schulte2012using},
%% evaluated at the two points.
In the approaches for the pair distribution function mentioned above,
the radial distribution function used to approximate pair correlation
was dependent upon the density averaged over some volume.  We seek to
achieve greater accuracy by using instead a function dependent upon
our averaged $g_{\sigma}(\rr)$ discussed above, which holds more
information about an inhomogeneous system than does a simple
convolution of the density.
%
We approximate the pair correlation with the average of two radial
distribution functions, evaluated at the distance between the two
points, that are themselves functions of the averaged pair correlation
at contact $g_{\sigma}(\rr)$ evaluated at the two points:
%
%% We therefore seek a function form
%% for our radial distribution that is dependent upon $g_{\sigma}(\rr)$
%% and that will approach the correct solution in the homogeneous limit
%% and also in the limit of contact distance between the two points.
\begin{align}
  g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) +
    g(|r_{12}|, g_\sigma(\rr_2))}{2}. \label{eq:g2-our-mean}
\end{align}
$g^{(2)}(\rr_1,\rr_2)$ above is constructed to reproduce the exact
value for the integral$ \int n(\rr_1)n(\rr_2)
g^{(2)}(\rr_1,\rr_2)\delta(|\rr_1-\rr_2|-\sigma)d\rr_1d\rr_2$.  \par
The approximation described in Eq.~\ref{eq:g2-our-mean} requires the
radial distribution function as a function of $r$ and $g_\sigma$.  One
approach to constructing $g(r,g_\sigma)$ is to compute $\eta$ from
$g_\sigma$ and then find $g(r,\eta)$ either using some approximate
functional form (e.g. Percus-Yevick) or using Monte Carlo simulation
results.  In this paper we first use the Monte Carlo results (which
are also used in evaluating the approximations of earlier works), and
refer to this approximation (which has an exact $g(r,g_\sigma)$) as
``MCRD.''  \jeffsays{I thought maybe for Monte Carlo Radial
  Distribution} We will then construct our own, more computationally
efficient radial distribution approximation, which is addressed in the
following section.

@@

%% The finalized form of our pair correlation function is:
%% \begin{align}
%%   g^{(2)}(\rr_1,\rr_2) = \frac{g(|r_{12}|, g_\sigma(\rr_1)) + g(|r_{12}|, g_\sigma(\rr_2))}{2}
%% \end{align}
%% \begin{figure}
%%   \centering
%%   \includegraphics[width=\columnwidth]{figs/ghs-g2}% \\
%%   %\includegraphics[width=\columnwidth]{figs/ghs-g-ghs}
%%   \caption{Plot of the radial distribution function, with our fit.}\label{fig:radial-distribution}
%% \end{figure}

%% In contrast to the previous works, which are based on the homogeneous
%% radial distribution function $g(r,\eta)$ expressed as a function of
%% distance and packing fraction, we have constructed our separable
%% approximation for the radial distribution function as $g(r,
%% g_\sigma)$, which is a function of the radius $r$ and its value at
%% contact $g_\sigma$.  One can convert between these two representation
%% using the Carnahan-Starling formula for $g_\sigma(\eta)$
%% \begin{equation}
%%   g_\sigma = \frac{1-\tfrac{\eta}{2}}{(1-\eta)^3}.
%% \end{equation}
%% The form $g(r, g_\sigma)$ is more convenient for treating the
%% inhomogeneous case, because we can directly compute the averaged value
%% of $g_\sigma$ for an inhomogeneous system~\cite{schulte2012using}.
\jeffsays{I just commented out a whole paragraph that it seems to me
  has become reduntant.  In it was the Carnahan-Starling
  approximation.  Do we need to put this back in?}

We construct $g(r, g_\sigma)$ to satisfy several constraints.  Naturally,
\begin{equation}
  g(\sigma, g_\sigma) = g_\sigma.
\end{equation}
A second constraint applies to the integral of the total correlation
function
\begin{align}
  1 + n\int h(r)d\rr &= nkT\chi_T \label{eq:total-constraint}
\end{align}
where $h(r) = g(r) - 1$ is the total correlation function.
A final constraint is that the slope of the radial distribution
function at contact is given to a very good approximation by
\begin{align}
  g'(\sigma) \approx - h_\sigma g_\sigma
\end{align}
where $h_\sigma = g_\sigma - 1$ is the total correlation function.
Under these three constraints, we applied a least-squares fit to
simulation data for the radial distribution function at packing
fractions from 0.05 to 0.4 by steps of 0.05.  Our fitted radial
distribution function is given by
\begin{multline}
  g(r,g_\sigma) = 1 + h_\sigma e^{-\kappaO \left(\frac{r}{\sigma}-1\right)} \\
  + h_\sigma(\kappaO - 2g_\sigma)  \left(\frac{r}{\sigma}-1\right)e^{-\kappaI  \left(\frac{r}{\sigma}-1\right)} \\
  + I(g_\sigma)  \left(\frac{r}{\sigma}-1\right)^2e^{-\kappaZ  \left(\frac{r}{\sigma}-1\right)}
\end{multline}
\begin{equation}
  I(g_\sigma) = \kappaZ^5 \frac{
    \frac{\chi-1}{24\eta} - h_\sigma \frac{(\kappaO-2
      g_\sigma)(\kappaI^2 + 4 \kappaI + 6)}{\kappaI^4}
    - h_\sigma\frac{\kappaO^2 + 2\kappaO + 2}{\kappaO}
  }{2 \kappaZ^2 + 12 \kappaZ + 24}
\end{equation}
\begin{equation}
  \chi = \frac{(1-\eta)^4}{\eta^4 - 4\eta^3 + 4\eta^2 + 4\eta + 1}
\end{equation}
\newcommand\nastyetacuberoot{\sqrt[3]{54 g_\sigma^2 -
    6\sqrt{81g_\sigma^4 - 6g_\sigma^3}}}
\begin{equation}
  \eta = 1 - \frac{1}{\nastyetacuberoot} - \frac{\nastyetacuberoot}{6g_\sigma}
\end{equation}
This function includes three fitted parameters, $\kappaO =
\kappazero$, $\kappaI = \kappaone$,
and $\kappaZ = \kappatwo$.
The fit is displayed in Fig.~\ref{fig:radial-distribution}.  In that
figure, it is apparent that the integral of $h(r)$ from
Eq.~\ref{eq:total-constraint} results in cancelling errors at large
distances.  The maximum error in $g(r)$ at packing fractions up to
0.4 is \maxerr, which occurs at $\eta = \etamaxerr$ and $r =
\rmaxerr$.

\newcommand\paircaption{\fixme{Look into smoothing the top plot on
    right.} The pair distribution function near a hard wall, with
  packing fractions of 0.1 and 0.3 and $\rr_1$ in contact with the
  hard wall.  On the left is a 2D plot of $g^{(2)}(\rr_1,\rr_2)$ as
  $\rr_2$ varies.  The top half of this figure shows the result of the
  Monte Carlo simulation, while the bottom half shows the result of
  the appproximation developed in this work.  On the right is a plot
  of $g^{(2)}(\rr_1,\rr_2)$ on the path illustrated in the figure to
  the left.  This plot compares the Monte Carlo results (black) with
  those of this work (blue), Sokolowski and Fischer
  (red)~\cite{sokolowski1992role}, and Fischer and Methfessel
  (green)~\cite{fischer1980born}.  The latter is only plotted at
  contact, where it is defined.  }

@@

We begin by examining the pair distribution function near a hard wall,
with a focus on the case where one of the two spheres is in contact
with the hard wall.  Figures~\ref{fig:pair-distribution}a,\ref{fig:pair-distribution}b
and Figures~\ref{fig:pair-distribution}c,\ref{fig:pair-distribution}d compare the results of our
approximate functional with Monte Carlo simulation at packing
fractions of 0.1 and 0.3 respectively.  We see reasonable agreement at
the lower density, with some reduced oscillations at larger distances,
and a flatter angular dependence when the two spheres are in contact.
At the higher density, we see significant structure developing in the
simulation results that is not reflected in our approximation.

Figures~\ref{fig:pair-distribution}b and~\ref{fig:pair-distribution}d
show the pair distribution function as plotted along paths
illustrated in Figures~\ref{fig:pair-distribution}a and
\ref{fig:pair-distribution}c.  These plots compare this work with
Monte Carlo results, as well as the approximations of Sokolowski and
Fischer~\cite{sokolowski1992role} and that of Fischer and
Methfessel~\cite{fischer1980born} at packing fractions of 0.1 and 0.3.
The approach of Fischer and Methfessel is only defined when the two
spheres are in contact, and is therefore only plotted on that segment
of the path.  As an input to both previous approximations we use the
hard sphere radial distribution function found with Monte Carlo
simulation, interpolated as necessary.  We find that both previous
approximations to the pair distribution function predict the
qualitative angular dependence of the pair distribution function at
contact better than this work.  However, in each case the pair
distribution function has a systematic error at contact---either too
high or too low.  Thus while our approximation is smoother than either
of the existing approaches, its errors will have a tendancy to cancel
when used in a perturbation expansion.  At higher densities, the
approximation of Fischer and Methfessel requires evaluating the radial
distribution function at densities significantly higher than the
freezing density, which poses numerical difficulties when using the
radial distribution function from simulation.  When the two points are
more than a radius away from contact, we find that any of these
approaches gives a reasonable prediction, with our approach
underestimating the oscillations in $g^{(2)}$, as expected based on
the fit in Fig.~\ref{fig:radial-distribution}.

@@

\begin{figure*}
  \includegraphics[width=\textwidth]{figs/triplet-correlation-pretty-contact-3.pdf}\\
  \includegraphics[width=\textwidth]{figs/triplet-correlation-pretty-inbetween-3.pdf}
  \caption{\fixme{Remove one of, or make a note of, the two curves for
      this work}\fixme{Fix MC issue here (volume wrong?)} The triplet
    distribution function $g^{(3)}(\rr_1,\rr_2,\rr_3)$ at packing
    fraction 0.3, plotted when $\rr_1$ and $\rr_2$ are in contact.
    The color plot on the left shows the triplet distribution
    function, with white corresponding to the contact value of the
    \emph{pair} distribution function.  The plot on the right shows
    the value of $g^{(3)}$ along the path shown as a dotted line on
    the plot to the left.}\label{fig:triplet-contact-distribution}
\end{figure*}

@@

A particularly relevant quantitative test of a pair distribution
function is how well it predicts the interaction energy due to a pair
potential.  To this end, we have computed the error in the first term
in a high-temperature perturbation expansion $a_1$ for several pair
potentials.  In order to focus on effects at the interface, we have
defined a position-dependent pair interaction energy as
\begin{align}
  \frac{dF_1}{dz} &=
  \tfrac12 \int g^{(2)}_{HS}(\rr,\rr')n(\rr)n(\rr')\Phi(|\rr-\rr'|)
  d\rr'\, dxdy\label{eq:da1}
\end{align}
which gives the contribution to the mean-field free energy due to
molecules located at position $z$.  \jeffsays{About this last
  sentence: this is molecules in the plane $z$, right?  Wanted to make
  sure that's the case before changing to something like 'in the
  plane'.  Saying the position z makes it seem that there's only one.}
